Datelife Reports

### 2019-09-30 biorxiv_2019.09.28.Rmd
for reproducibility's sake

---
output:
  bookdown::pdf_document2:
    keep_tex: yes
    toc: false
    includes:
      # in_header: preamble.tex
      after_body: [biorxiv_2019.09.28_figures.tex]
    # fig_caption: no
textalign: left  # this does not work...
geometry: margin = 1in
bibliography: library_red.bib
csl: systematic-biology.csl
link-citations: yes
# indent: true # this works
# header-includes:
#    - \setlength\parindent{24pt}  # this works too, but I'm gonna indent manually with 4 spaces
# nope, 4 spaces indents are not recognized, they do something else
linestretch: 2
header-includes:
    - \usepackage{caption}
    - \usepackage{blindtext}
    - \usepackage[left]{lineno}
    - \linenumbers
  # - \usepackage{xr}
    # - \usepackage[utf8]{inputenc}
    # - \usepackage[english]{babel}
    # - \usepackage[document]{ragged2e}
nocite: |
  @barker2012going, @barker2015new, @burns2014phylogenetics, @claramunt2015new, @gibb2015new, @Hedges2015, @hooper2017chromosomal, @Jetz2012, @price2014niche
---
<!-- \begin{\nolinenumbers} -->
<!-- Running head: DATELIFE: REVEALING THE DATED TREE OF LIFE <!--MAX 50 CHARACTERS, ALL CAPITALS-->

>>>>

---
output:
  bookdown::pdf_document2:
    keep_tex: yes
    toc: false
    includes:
      # in_header: preamble.tex
      after_body: [biorxiv_2019.09.28figcaps.tex]
    # fig_caption: no
textalign: left  # this does not work...
geometry: margin = 1in
bibliography: paper_references.bib
csl: systematic-biology.csl
link-citations: yes
# indent: true # this works
# header-includes:
#    - \setlength\parindent{24pt}  # this works too, but I'm gonna indent manually with 4 spaces
# nope, 4 spaces indents are not recognized, they do something else
linestretch: 2
header-includes:
    - \usepackage{caption}
    - \usepackage{blindtext}
    - \usepackage[left]{lineno}
    - \linenumbers
  # - \usepackage{xr}
    # - \usepackage[utf8]{inputenc}
    # - \usepackage[english]{babel}
    # - \usepackage[document]{ragged2e}
nocite: |
  @barker2012going, @barker2015new, @burns2014phylogenetics, @claramunt2015new, @gibb2015new, @Hedges2015, @hooper2017chromosomal, @Jetz2012, @price2014niche
---
<!-- \begin{\nolinenumbers} -->
<!-- Running head: DATELIFE: REVEALING THE DATED TREE OF LIFE <!--MAX 50 CHARACTERS, ALL CAPITALS-->

### 2019-09-30 biorxiv_2019.09.28.tex
for reproducibility's sake

A tool for efficient reuse of expert, published data on time of lineage divergence should have an open and fuly public chronogram database storing data in a format suitable for scientific reuse, an automatised way of accessing the information, and straightforward means of comparing and summarizing chronogram information as needed by the user.
A prototype service aiming to meet this characteristics was developed over a series of hackathons at the National Evolutionary Synthesis Center (Stoltzfus et al. \protect\hyperlink{ref-Stoltzfus2013}{2013}).
In here we present the formal description and implementation of the \texttt{datelife} service, constituted by an R package and a web site (\url{http://www.datelife.org/}). There is still much room for improvement, and flaws and limitations are addressed below. We strived for the current implementation of \texttt{datelife} to perform the basic tasks described above, featuring a system for maintenance of an open database of chronograms pulled from public repositories, methods to summarize and compare source chronograms, and new functions to visualize and graphically compare source and summary chronograms.

@@

\texttt{datelife} is free and open source and it can be used through its current website
\url{http://www.datelife.org/query/}, through its R package, and through Phylotastic's project web portal \url{http://phylo.cs.nmsu.edu:3000/}.
\texttt{datelife}'s website is maintained using RStudio's shiny server and the shiny package open infrastructure, as well as Docker.
\texttt{datelife}'s R package stable version is \textbf{not} available yet
for installation from the CRAN repository (\url{https://cran.r-project.org/package=datelife})
using the command \texttt{install.packages(pkgs\ =\ "datelife")} from within R. Development versions
are available from the GitHub repository (\url{https://github.com/phylotastic/datelife})
and can be installed using the command \texttt{devtools::install\_github("phylotastic/datelife")}.

@@

We thank colleagues from the O'Meara Lab at the University
of Tennesse Knoxville for suggestions, discussions and software testing.
The late National Evolutionary Synthesis Center (NESCent), which sponsored hackathons
that led to initial work on this project. The team that assembled datelife's first proof of concept: Tracy Heath, Jonathan Eastman, Peter Midford, Joseph Brown, Matt Pennell, Mike Alfaro, and Luke Harmon.
The Open Tree of Life project that provides the open, metadata rich repository of
trees used for \texttt{datelife}.
The many scientists who publish their chronograms in an open, reusable form, and
the scientists who curate them for deposition in the Open Tree of Life repository.
The NSF for funding nearly all the above, in addition to the ABI grant that funded this project itself.

\newpage
\begin{center}
\textsc{References}
\end{center}

@@

\newpage
\begin{figure}[!h]
\includegraphics{Fig1.pdf}
\caption{}
\label{fig:workflow}
\end{figure}
% \begin{center}
% \textsc{Figure \ref{fig:workflow}}
% \end{center}
Stylized DateLife workflow. This shows the general worflows and analyses that can be performed with \texttt{datelife}, via the R package or through the website. Details on the functions involved on each workflow are shown in \texttt{datelife}'s R package vignette.
\newpage
\begin{figure}[!h]
\includegraphics[width=1\linewidth]{fig_runtime_main.pdf}
\caption{}
\label{fig:runtime1}
\end{figure}
% \begin{center}
% \textsc{Figure \ref{fig:runtime1}}
% \end{center}
Computation time of query processing and search across \texttt{datelife}'s chronogram database relative to number of input taxon names. We sampled N names from the class Aves for each cohort 100 times and then performed a search with query processing not using the Taxon Names Resoultion Service (TNRS; dark gray), and using TNRS (light gray). We also performed a search using the already processed query for comparison (light blue). 
\newpage
\begin{figure}[!h]
\includegraphics[width=1\linewidth]{fig_schronograms1.pdf}
\caption{}
\label{fig:schronograms1}
\end{figure}
% \begin{center}
% \textsc{Figure \ref{fig:schronograms1}}
% \end{center}
Lineage through time (LTT) plots of source chronograms containing all or a subset of species from the bird family Fringillidae of true finches. Arrows indicate maximum age of each chronogram. Numbers reference to chronograms' original publications 1: Barker et al. (\protect\hyperlink{ref-barker2012going}{2012}), 2: Barker et al. (\protect\hyperlink{ref-barker2015new}{2015}), 3: Burns et al. (\protect\hyperlink{ref-burns2014phylogenetics}{2014}), 4: Claramunt and Cracraft (\protect\hyperlink{ref-claramunt2015new}{2015}), 5: Gibb et al. (\protect\hyperlink{ref-gibb2015new}{2015}), 6: Hedges et al. (\protect\hyperlink{ref-Hedges2015}{2015}), 7: Hooper and Price (\protect\hyperlink{ref-hooper2017chromosomal}{2017}), 8: Jetz et al. (\protect\hyperlink{ref-Jetz2012}{2012}), 9: Price et al. (\protect\hyperlink{ref-price2014niche}{2014}).
\newpage
\begin{figure}[!h]
\includegraphics{fig_summaries.pdf}
%\includegraphics{fig_summaries.pdf}
\caption{}
\label{fig:summaries}
\end{figure}
% \begin{center}
% \textsc{Figure \ref{fig:summaries}}
% \end{center}
LTT plots of median and Supermatrix Distance Method (SDM) chronograms summarising information from source chronograms found for the Fringillidae. Arrows indicate maximum age.
\newpage
\begin{figure}[!h]
\includegraphics{fig_crossval_bladj.pdf}
\caption{}
\label{fig:cvbladj}
\end{figure}
% \begin{center}
% \textsc{Figure \ref{fig:cvbladj}}
% \end{center}
LTT plots showing results from the cross-validation analyses of trees without branch lengths dated using BLADJ. The dating analysis can only be performed in trees with more than 2 tips, thus excluding chronogram from study 4; its data was still used as calibration for the other source chronograms.
\newpage
\begin{figure}[!h]
\includegraphics{fig_crossval_boldsumm.pdf}
\caption{}
\label{fig:cvbold}
\end{figure}
% \begin{center}
% \textsc{Figure \ref{fig:cvbold}}
% \end{center}
LTT plots showing results from the cross-validation analyses of trees with branch length reconstructed with data from the Barcode of Life Database (BOLD) dated using PATHd8. We could construct a tree with branch lengths for all source chronograms. However, dating with PATHd8 was only successful in three source chronograms shown here.

>>>>

A tool for efficient reuse of expert, published data on time of lineage divergence should have an open and fully public chronogram database storing data in a format suitable for scientific reuse, an automatised way of accessing the information, and straightforward means of comparing and summarizing chronogram information as needed by the user.
A prototype service aiming to meet this characteristics was developed over a series of hackathons at the National Evolutionary Synthesis Center (Stoltzfus et al. \protect\hyperlink{ref-Stoltzfus2013}{2013}).
In here we present the formal description and implementation of the \texttt{datelife} service, constituted by an R package and a web site (\url{http://www.datelife.org/}). There is still much room for improvement, and flaws and limitations are addressed below. We strived for the current implementation of \texttt{datelife} to perform the basic tasks described above, featuring a system for maintenance of an open database of chronograms pulled from public repositories, methods to summarize and compare source chronograms, and new functions to visualize and graphically compare source and summary chronograms.

@@

\texttt{datelife} is free and open source and it can be used through its current website
\url{http://www.datelife.org/query/}, through its R package, and through Phylotastic's project web portal \url{http://phylo.cs.nmsu.edu:3000/}.
\texttt{datelife}'s website is maintained using RStudio's shiny server and the shiny package open infrastructure, as well as Docker.
\texttt{datelife}'s R package stable version will be available
for installation from the CRAN repository (\url{https://cran.r-project.org/package=datelife})
using the command \texttt{install.packages(pkgs\ =\ "datelife")} from within R. Development versions
are available from the GitHub repository (\url{https://github.com/phylotastic/datelife})
and can be installed using the command \texttt{devtools::install\_github("phylotastic/datelife")}.

@@

We thank colleagues from the O'Meara Lab at the University
of Tennesse Knoxville for suggestions, discussions and software testing.
The late National Evolutionary Synthesis Center (NESCent), which sponsored hackathons
that led to initial work on this project. The team that assembled \texttt{datelife}'s first proof of concept: Tracy Heath, Jonathan Eastman, Peter Midford, Joseph Brown, Matt Pennell, Mike Alfaro, and Luke Harmon.
The Open Tree of Life project that provides the open, metadata rich repository of
trees used for \texttt{datelife}.
The many scientists who publish their chronograms in an open, reusable form, and
the scientists who curate them for deposition in the Open Tree of Life repository.
The NSF for funding nearly all the above, in addition to the ABI grant that funded this project itself.
\newpage

\begin{center}
\textsc{References}
\end{center}

@@

\newpage
\begin{figure}[!h]
\includegraphics{./figures/Fig1.pdf}
\caption{}
\label{fig:workflow}
\end{figure}
% \begin{center}
% \textsc{Figure \ref{fig:workflow}}
% \end{center}
Stylized DateLife workflow. This shows the general worflows and analyses that can be performed with \texttt{datelife}, via the R package or through the website. Details on the functions involved on each workflow are shown in \texttt{datelife}'s R package vignette.
\newpage
\begin{figure}[!h]
\includegraphics[width=1\linewidth]{./figures/fig_runtime_main.pdf}
\caption{}
\label{fig:runtime1}
\end{figure}
% \begin{center}
% \textsc{Figure \ref{fig:runtime1}}
% \end{center}
Computation time of query processing and search across \texttt{datelife}'s chronogram database relative to number of input taxon names. We sampled N names from the class Aves for each cohort 100 times and then performed a search with query processing not using the Taxon Names Resoultion Service (TNRS; dark gray), and using TNRS (light gray). We also performed a search using the already processed query for comparison (light blue).
\newpage
\begin{figure}[!h]
\includegraphics[width=1\linewidth]{./figures/fig_schronograms1.pdf}
\caption{}
\label{fig:schronograms1}
\end{figure}
% \begin{center}
% \textsc{Figure \ref{fig:schronograms1}}
% \end{center}
Lineage through time (LTT) plots of source chronograms containing all or a subset of species from the bird family Fringillidae of true finches. Arrows indicate maximum age of each chronogram. Numbers reference to chronograms' original publications 1: Barker et al. (\protect\hyperlink{ref-barker2012going}{2012}), 2: Barker et al. (\protect\hyperlink{ref-barker2015new}{2015}), 3: Burns et al. (\protect\hyperlink{ref-burns2014phylogenetics}{2014}), 4: Claramunt and Cracraft (\protect\hyperlink{ref-claramunt2015new}{2015}), 5: Gibb et al. (\protect\hyperlink{ref-gibb2015new}{2015}), 6: Hedges et al. (\protect\hyperlink{ref-Hedges2015}{2015}), 7: Hooper and Price (\protect\hyperlink{ref-hooper2017chromosomal}{2017}), 8: Jetz et al. (\protect\hyperlink{ref-Jetz2012}{2012}), 9: Price et al. (\protect\hyperlink{ref-price2014niche}{2014}).
\newpage
\begin{figure}[!h]
\includegraphics{./figures/fig_summaries.pdf}
%\includegraphics{fig_summaries.pdf}
\caption{}
\label{fig:summaries}
\end{figure}
% \begin{center}
% \textsc{Figure \ref{fig:summaries}}
% \end{center}
LTT plots of median (top) and Supermatrix Distance Method (SDM; bottom) chronograms summarising information from source chronograms found for the Fringillidae. Arrows indicate tree maximum age.
\newpage
\begin{figure}[!h]
\includegraphics{./figures/fig_crossval_bladj.pdf}
\caption{}
\label{fig:cvbladj}
\end{figure}
% \begin{center}
% \textsc{Figure \ref{fig:cvbladj}}
% \end{center}
LTT plots showing results from the cross-validation analyses of trees without branch lengths dated using BLADJ. The dating analysis can only be performed in trees with more than 2 tips, thus excluding chronogram from study 4; its data was still used as calibration for the other source chronograms.
\newpage
\begin{figure}[!h]
\includegraphics{./figures/fig_crossval_boldsumm.pdf}
\caption{}
\label{fig:cvbold}
\end{figure}
% \begin{center}
% \textsc{Figure \ref{fig:cvbold}}
% \end{center}
LTT plots showing results from the cross-validation analyses of trees with branch length reconstructed with data from the Barcode of Life Database (BOLD) dated using PATHd8. We could construct a tree with branch lengths for all source chronograms. However, dating with PATHd8 was only successful in three source chronograms shown here.

### 2019-09-30 paper.Rmd
for reproducibility's sake

```{r eval = TRUE, echo = FALSE, warning=FALSE, message=FALSE, results='hide'}
# code to generate the paper data and figures:
# source("data.R")
# inactivated now just for testing purposes
# we are using drake to avoid running the examples everytime we change something in the Rmd and knit
# using this for testing instead:
source("./frin_plan.R") # stopped using drake for this, it is better to run the analysis directly
loadd() # loads all the targets listed in the workflow plan of the previous `make()`
library(datelife)
data(opentree_chronograms)
rpackage_available <- FALSE
availability <- ifelse(rpackage_available, "is available", "is **not** available yet")
thank_origteam <- TRUE
origteam <- ifelse(thank_origteam, "The team that assembled datelife's first proof of concept: Tracy Heath, Jonathan Eastman, Peter Midford, Joseph Brown, Matt Pennell, Mike Alfaro, and Luke Harmon.", "")
```

@@

We thank colleagues from the O'Meara Lab at the University
of Tennesse Knoxville for suggestions, discussions and software testing.
The late National Evolutionary Synthesis Center (NESCent), which sponsored hackathons
that led to initial work on this project. `r origteam`
The Open Tree of Life project that provides the open, metadata rich repository of
trees used for `datelife`.
The many scientists who publish their chronograms in an open, reusable form, and
the scientists who curate them for deposition in the Open Tree of Life repository.
The NSF for funding nearly all the above, in addition to the ABI grant that funded this project itself.

>>>>

---
output:
  pdf_document: default
  html_document: default
---
```{r eval = TRUE, echo = FALSE, warning=FALSE, message=FALSE, results='hide'}
# code to generate the paper data and figures:
# source("data.R")
# inactivated now just for testing purposes
# we are using drake to avoid running the examples everytime we change something in the Rmd and knit
# using this for testing instead:
source("./data/frin_plan.R") # stopped using drake for this, it is better to run the analysis directly
loadd() # loads all the targets listed in the workflow plan of the previous `make()`
library(datelife)
data(opentree_chronograms)
rpackage_available <- FALSE
availability <- ifelse(rpackage_available, "is available", "will be available")
```

@@

We thank colleagues from the O'Meara Lab at the University
of Tennesse Knoxville for suggestions, discussions and software testing.
The late National Evolutionary Synthesis Center (NESCent), which sponsored hackathons
that led to initial work on this project. The team that assembled `datelife`'s first proof of concept: Tracy Heath, Jonathan Eastman, Peter Midford, Joseph Brown, Matt Pennell, Mike Alfaro, and Luke Harmon.
The Open Tree of Life project that provides the open, metadata rich repository of
trees used for `datelife`.
The many scientists who publish their chronograms in an open, reusable form, and
the scientists who curate them for deposition in the Open Tree of Life repository.
The NSF for funding nearly all the above, in addition to the ABI grant that funded this project itself.

### 2019-09-30 paper_drafts/Datelife_paper1.Rmd
for reproducibility's sake

```{r eval = TRUE, echo = FALSE, warning=FALSE, message=FALSE, results='hide'}
source("Datelife_paper1_global.R")
```

@@

\newpage
Clade ages represent a fundamental piece of information for evolutionary
understanding in many areas of research, from developmental to conservation biology [@Felsenstein1985a; @Webb2000], from historical biogeography to species diversification studies [@posadas2006historical; @Morlon2014]. The primary information needed these time estimates comes from the fossil record. Coupled with phylogenies with branch lengths based on molecular and/or morphological data, the time of divergence of extant and extinct lineages can be reconstructed with molecular dating methods.
The number of studies publishing phylogenies with branch lengths proportional to geological time (hereafter chronograms) have constantly increased in number for the last two decades [@Kumar2017].
Still, generating a chronogram is not an easy task unless you have specialized training: it requires inferring a tree, understanding what fossil data are available and their limits, and where fossils go on the tree. That is why there has been an urge for promoting and facilitating reuse of the vast amount of phylogenetic and time of lineage divergence data that has been generated and made available in publications, for the advantage of research relying on this information [@webb2005phylomatic; @Stoltzfus2013].

>>>>

```{r eval = TRUE, echo = FALSE, warning=FALSE, message=FALSE, results='hide'}
# code to generate the paper data and figures:
source("Datelife_paper1_global.R")
```

@@

\newpage
Clade ages represent a fundamental piece of information for evolutionary
understanding in many areas of research, from developmental to conservation biology [@Felsenstein1985a; @Webb2000], from historical biogeography to species diversification studies [@posadas2006historical; @Morlon2014]. The primary information needed for these time estimates comes from the fossil record. Coupled with phylogenies with branch lengths based on molecular and/or morphological data, the time of divergence of extant and extinct lineages can be reconstructed with molecular dating methods.
The number of studies publishing phylogenies with branch lengths proportional to geological time (hereafter chronograms) have constantly increased in number for the last two decades [@Kumar2017].
Still, generating a chronogram is not an easy task unless you have specialized training: it requires inferring a tree, understanding what fossil data are available and their limits, and where fossils go on the tree. That is why there has been an urge for promoting and facilitating reuse of the vast amount of phylogenetic and time of lineage divergence data that has been generated and made available in publications, for the advantage of research relying on this information [@webb2005phylomatic; @Stoltzfus2013].

### 2019-09-30 paper_drafts/Datelife_paper1_figures.Rmd
for reproducibility's sake

---
output: 
  bookdown::pdf_document2:
    keep_tex: yes
textalign: left
geometry: margin = 1in
bibliography: library_red.bib
csl: systematic-biology.csl
link-citations: yes
linestretch: 2
header-includes:
  - \usepackage[left]{lineno}
  - \linenumbers
  - \usepackage{caption}
  # - \usepackage{xr}
  #- \externaldocument{name of external document to reference fugures from there in here}
---

@@

\begin{figure}[!h]
\includegraphics{fig_runtime1.pdf}
\caption{}
\label{fig:runtime1}
\end{figure}

>>>>

---
output: 
  bookdown::pdf_document2:
    keep_tex: yes
textalign: left
geometry: margin = 1in
bibliography: library_red.bib
csl: systematic-biology.csl
link-citations: yes
linestretch: 2
header-includes:
  - \usepackage[left]{lineno}
  - \linenumbers
  # - \usepackage{caption} # no need to use this one here I think, use it on the yaml of the general file
  # - \usepackage{xr}
  #- \externaldocument{name of external document to reference fugures from there in here}
---

@@

\begin{figure}[!h]
\includegraphics{fig_runtime_main.pdf}
\caption{}
\label{fig:runtime1}
\end{figure}
